Efficient and Effective Metric Space Indexing for Complete Record Linkage Al: Exploring approaches to probabilistic record linkage using similarity searching (metric space indexing?)

### 2017-11-12 paper.tex
Update on Overleaf.

something about record linkage and its importance, application examples

drawback of existing approaches: blocking and comparisons/classification
done separately, sometimes if bad linkage quality means to go back to redo
blocking.

something about similarity spaces / metric spaces, M-trees and its
more recent advanced/improved versions, and how they have been used in
similarity search.

How our work is different from these other metric space works, and
description of our contributions

The motivation for this work is the Digitising Scotland project (Dibben 2012), which is in the process of transcribing and linking all the vital events recorded in Scotland between 1856 and 1977. This data set will, when complete, include around 14 million birth records, 11 million death records and 4 million marriage records. As part of the work, certain data fields (locations, occupations and causes of death) are also being classified to the relevant standard coding schemes.

@@

summarise related work in blocking/indexing

>>>>

% Peter, 20171112:

Record linkage, also known as entity resolution, data matching and
duplicate detection~\cite{Chr12}, is the process of identifying and
matching records that refer to the same real-world entities within or
across databases. The entities to be linked are often people (such as 
patients in hospital or customers in business databases), but record
linkage can also be applied to link consumer products or bibliographic
records~\cite{Chr12}. Record linkage is commonly challenged by the
lack of unique entity identifiers (keys) in the databases to be linked,
which prevents the use of a database join. Instead, the linkage of
records requires the comparison of the attributes (or fields) that are
available in common in the databases to be linked. For databases that
contain information about individuals, these attributes include the
names, address, dates of birth, and so on, of individuals.

To overcome data quality issues such as typographical errors and
variations (which are common in name and address values~\cite{Chr12}),
approximate string comparison functions (such as edit distance, the
Jaro-Winkler comparator, or Jaccard similarity~\cite{Chr12}) are used
to compare pairs of records, leading to a vector of similarities (one
similarity per attribute/field compared) for each pair. These
similarity vectors are then used to classify the compared record
pairs into matches (where it is assumed both records in a pair
correspond to the same real-world entity) and non-matches (where the
records are assumed to correspond to two different entities). Various
classification methods have been employed in record
linkage~\cite{Chr12,Don15}, ranging from simple threshold based to
sophisticated clustering and supervised classification techniques, as
well as active learning approaches~\cite{Wan15}.

Besides the two challenges of a lack of unique entity identifiers and
data quality (which will affect linkage quality), record linkage is
also challenged by the increasing sizes of the databases to be
linked~\cite{Don15}. To prevent the full pair-wise comparison of all
possible record pairs (which is quadratic in the sizes of the
databases to be linked), blocking techniques (commonly known as
\emph{indexing}~\cite{Chr12b}) are used. The basic idea of these
techniques is to split the databases into smaller blocks in a
computationally efficient way such that records that likely correspond
to the same entity are inserted into the same block. Only records that
are in the same block are then compared with each other in more
details.

While indexing techniques facilitate the efficient linkage of very
large data-bases~\cite{Don15}, they generally achieve scalability at
the cost of loss in linkage quality, because potentially true matching
record pairs are removed in the indexing step, leading to a reduction
in recall of the final linkage result~\cite{Chr12}. A variety of
indexing techniques, as discussed in more details in the following
section, have been proposed, ranging from simple phonetic based
blocking~\cite{Chr12} and sorting of the databases~\cite{Dra12} to
locality sensitive hashing based techniques~\cite{Kim10,Steorts2014}.
Unsupervised~\cite{Kej13,Ram15} and supervised~\cite{Bil06,Mic06} 
learning techniques of optimal blocking schemes have also been
proposed.

\begin{figure}[!t]
  \centering
  %\scalebox{1.0}[1.0]
  \includegraphics[width=1.0\textwidth]{figures/linkage-process}
  \caption{Overview of the steps of the traditional record linkage
           process (left-hand size) and our proposed metric space
           indexing approach (right-hand size), as described in
           Sect.~\ref{sec-intro}, where records from two databases,
           $\mathbf{D}_A$ and $\mathbf{D}_B$, are being linked.}
           \label{fig-rl-process}
\end{figure}

The overall process of a record linkage application, where indexing is
preformed prior to the comparison and classification steps, as
illustrated in Fig.~\ref{fig-rl-process}, adds a further practical
complexity. Indexing, comparison and classification are often conducted
using algorithms selected (and their parameters set) based on technical
and domain expertise of the record linkage user, followed by a manual
assessment of the linkage outcomes~\cite{Chr12}. If the quality of the
resulting links is not good enough for a certain application, the
linkage process needs to be repeated with different parameter settings
and possibly also alternative algorithms selected. This can result in
a time-consuming and labor intensive iterative process. A major
challenge is often the heuristic nature of indexing, where the choice
of a certain indexing technique and its parameters (including which
attributes to use in indexing) will determine the final outcome of a
linkage.
% Repeat:
% If a certain blocking approach results in a loss of recall in the
%final linkage results, the blocking needs to be redone, an often time
%consuming process.

In this paper we explore how existing indexing techniques can lead to
loss in recall, even for record pairs that are highly similar and
likely correspond to true matches. As a result, using such techniques
can lead to \emph{incomplete} record linkage results. To overcome this
problem, we advocate the use of \emph{metric space indexing} (MSI)
approaches, where it is possible to guarantee that all record pairs,
up-to a certain distance threshold, are included in the record linkage
results. MSI also prevents the expensive detailed comparison of record
pairs that have low similarities but that need to be compared because
they have been inserted into the same block with traditional indexing
techniques. Furthermore, MSI allows us to combine the indexing,
comparison and classification steps (as traditionally conducted
independently) into one single step, as shown in
Fig.~\ref{fig-rl-process}, making the overall record linkage process
both more efficient and more effective.

% Peter 20171112: do we also need to mention we do not waste
% comparisons of record pairs with low similarities (i.e. those with
% a distance > max_dist)?

~ \\
\emph{Peter 20171111: here we need the figure showing that traditional
blocking techniques result in incomplete record linkage, while metric
indexing does not.}
~ \\

While MSI has been used extensively in areas such as database indexing
and general nearest neighbor search, its use in record linkage so far
has been limited. 
% Peter 20171112: a pretty dangerous statement. if wrong we're stuffed.. 
% but I am not really aware of MSI approaches in RL..
Many different MSI approaches have been developed, such as various
metric tree structures~\cite{}. 
% Peter 20171112: needs same text from you guys on MSI..
In this paper, we will use the MSI techniques of M-tree and M-file (?)
and show how their use can facilitate an improvement of both
efficiency and effectiveness of the record linkage process.

~ \\
\emph{
to add: something about similarity spaces / metric spaces, M-trees and its
more recent advanced/improved versions, and how they have been used in
similarity search.}

\emph{to add: How our work is different from these other metric space works, and
description of our contributions.
}
~ \\

\smallskip
\textbf{Contributions:} We make three specific contributions in this
paper: (1) We investigate how existing indexing techniques can result
in \emph{incomplete} record linkage results, as well as in an
in-efficient record linkage process. (2) We explore MSI for record
linkage and propose a MSI based linkage approach that achieves
\emph{complete}, efficient and effective record linkage. (3) We
evaluate our approach on several real-world data sets from diverse
domains and show its advantages over existing indexing techniques for
record linkage.

% old stuff below

\iffalse  % *******************************************************
% we can't write this as it reveals us in the double-blind review
% we could ad to final paper
The motivation for this work is the Digitising Scotland project (Dibben 2012), which is in the process of transcribing and linking all the vital events recorded in Scotland between 1856 and 1977. This data set will, when complete, include around 14 million birth records, 11 million death records and 4 million marriage records. As part of the work, certain data fields (locations, occupations and causes of death) are also being classified to the relevant standard coding schemes.
\fi  % *********************************************************

@@

We now briefly review relevant related work in the areas of indexing
for record linkage (for a recent survey see Christen~\cite{Chr12b}),
and metric space indexing.

\smallskip
\textbf{Indexing for Record Linkage:}
Techniques to link records across databases have been developed over
the past five decades~\cite{Fel69}, and scaling the linkage of large
databases has been an ongoing challenge as databases are continue to
grow in size and complexity. Traditional blocking techniques use a
single or a combination of attributes (known as a \emph{blocking key})
to insert records that have the same value(s) in such a blocking key
into the same block. Only records in the same block are then combined
with each other. To overcome variations and misspellings, the values
used in blocking keys are commonly phonetically encoded using functions
such as Soundex, NYSSI, or Double-Metaphone~\cite{Chr12}. These
techniques convert a string (assumed to be a name) into a code
according to how the name sounds, assigning the same code to similar
sounding names (such as `Gail' and `Gayle' or `Dickson' and `Dixon').

A different approach to indexing is the sorted neighborhood 
method~\cite{Mon96}, where the databases to be linked are sorted
according to a \emph{sorting key} (usually a concatenation of the
values from several attributes). A sliding window is then moved over
the databases and only records within the same window are being
compared. A fixed window size, $w$, leads to a number of record
pair comparison of $O(n \times w)$, where $n = |\mathbf{D}_A| +
|\mathbf{D}_b|$, is the number of records in the two databases
$\mathbf{D}_A$ and $\mathbf{D}_B$ to be linked. Techniques that
adaptively shrink or expand the window size based on the
characteristics of the sorting key values have shown to improve both
linkage efficiency and quality~\cite{Dra12}. The drawback of sorting
based

\emph{peter 20171112 until here, to do: LSH Harra, Kim10, Stoerts, drawbacks of exiting techniques}
blocking framework papadakis TKDE 2013

% describe LSH: harra (Kim10)
% Stoerts

% learning blocking keys

Drawbacks of existing techniques: errors in blocking/sorting keys,
manually selected, is learned need manual training data, or are
expensive (if unsupervised)

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

\smallskip
\textbf{Metric Space Indexing:}

% Al's writing below
